To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED

### 2023-10-06 manuscript/manuscript.Rmd
aking updated to manuscript

---
title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization. A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity. Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes. The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

# Introduction {#sec:1}

@@

We start by evaluating the reduced-form effects of ED provider batch ordering tendency on downstream patient outcomes and turnaround time. We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences. Being treated by a provider in the top decile of the leniency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].

Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes. This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and hospital effeciency. It is important to note that this paper studies the impact of an batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.

The remainder of this paper is structured as follows. The next section describes the data source and outlines our baseline sample. The empirical strategy and its accompanying identifying assumptions are laid out in Section II. Section III presents the results. Section IV draws implications for batch-ordering policies. The last section concludes.

# Data and Definitions {#sec:2}

@@

We identified three distinct forms of test batching:

- Lab/Image Batch: The first test ordered for a patient was either an imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an additional imaging or lab test order within 5 minutes.

- Image/Image Batch: The first test ordered for a patient was an imaging test (CT scan, X-ray, Ultrasound), followed by an additional imaging test order within 5 minutes.

- Any Batch: This refers to instances where either a Lab/Image Batch or an Image/Image Batch occurred.

We conducted a sensitivity analysis for this definition, documented in Appendix 2. Table 1 highlights the considerable variation in patient complexity between those who had their tests batch ordered versus those where tests were ordered sequentially.

@@

Table 1 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered either individually (Sequential) or as part of a group (Batch) during an ED encounter from our baseline sample. Around 43% of patient encounters involved diagnostic tests ordered as part of a batch, the majority of which (60%) included two tests.

Differences were noted between the two cohorts. Patients who had batched tests ordered tended to have a longer ED length of stay, a lower Emergency Severity Index (ESI), and were generally older than those who had their tests ordered individually. Patients in the batch order group were also more likely to present with symptoms at triage, including tachycardia, tachypnea, fever, and hypotension. A deeper dive into the ESI data reveals a higher proportion of patients with more urgent conditions (ESI 1 and 2) in the batched group than in the sequential group. This finding might be indicative of the complexity and severity of the conditions of patients who necessitate batch testing.

Age distributions between the two groups demonstrate a greater proportion of patients aged 65 and above in the batch group, suggesting that older patients are more likely to have multiple tests ordered at once, possibly due to the presence of multiple comorbidities. The batched group showed a longer average ED length of stay compared to the sequential group. This extended stay may be reflective of the additional time required to conduct and evaluate the results of multiple tests. Symptoms presentation at triage also differed between the two groups, with the batched group being more likely to present with tachypnea, fever, and hypotension. However, tachycardia was found to be similar across both groups.

\begin{table}[htbp]
\centering
\tiny
\caption{Summary Statistics for Sequential vs. Batched Test Groups}
\begin{tabular}{lcccc}
\hline
Variable & \multicolumn{1}{c}{Overall Visits, N = 41,197\textsuperscript{1}} & \multicolumn{1}{c}{Sequential Arm, N = 23,4331} & \multicolumn{1}{c}{Batching Arm, N = 17,7641} & \multicolumn{1}{c}{p-value\textsuperscript{2}} \\
\hline
\textbf{ESI} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 474 (1.2\%) & 130 (0.6\%) & 344 (1.9\%) & \\
\hspace{0.5cm} 2 & 13,911 (34\%) & 6,502 (28\%) & 7,409 (42\%) & \\
\hspace{0.5cm} 3 & 23,655 (57\%) & 14,350 (61\%) & 9,305 (52\%) & \\
\hspace{0.5cm} 4 & 3,123 (7.6\%) & 2,422 (10\%) & 701 (3.9\%) & \\
\hspace{0.5cm} 5 & 34 ($<$0.1\%) & 29 (0.1\%) & 5 ($<$0.1\%) & \\
\hline
\textbf{Age} & & & & $<$0.001 \\
\hspace{0.5cm} $<$20 & 790 (1.9\%) & 605 (2.6\%) & 185 (1.0\%) & \\
\hspace{0.5cm} 20-45 & 9,412 (23\%) & 6,242 (27\%) & 3,170 (18\%) & \\
\hspace{0.5cm} 45-65 & 12,826 (31\%) & 7,393 (32\%) & 5,433 (31\%) & \\
\hspace{0.5cm} 65+ & 18,169 (44\%) & 9,193 (39\%) & 8,976 (51\%) & \\
\hline
\textbf{ED LOS} & & & & $<$0.001 \\
\hspace{0.5cm} Mean (SD) & 270 (382) & 258 (177) & 286 (545) & \\
\hline
\textbf{72hr Return} & 1,482 (3.6\%) & 957 (4.1\%) & 525 (3.0\%) & $<$0.001 \\
\hline
\textbf{Number of Tests} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 12,961 (31\%) & 11,937 (51\%) & 1,024 (5.8\%) & \\
\hspace{0.5cm} 2 & 19,112 (46\%) & 8,457 (36\%) & 10,655 (60\%) & \\
\hspace{0.5cm} 3 & 7,688 (19\%) & 2,581 (11\%) & 5,107 (29\%) & \\
\hspace{0.5cm} 4 & 1,342 (3.3\%) & 432 (1.8\%) & 910 (5.1\%) & \\
\hspace{0.5cm} 5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\hline
\textbf{Disposition} & & & & $<$0.001 \\
\hspace{0.5cm} Discharge & 25,302 (61\%) & 15,922 (68\%) & 9,380 (53\%) & \\
\hspace{0.5cm} Admit & 9,097 (22\%) & 4,209 (18\%) & 4,888 (28\%) & \\
\hspace{0.5cm} Other & 6,798 (17\%) & 3,302 (14\%) & 3,496 (20\%) & \\
\hline
\textbf{Tachycardic} & 8,115 (20\%) & 4,562 (19\%) & 3,553 (20\%) & 0.18 \\
\hline
\textbf{Tachypneic} & 3,822 (9.3\%) & 1,669 (7.1\%) & 2,153 (12\%) & $<$0.001 \\
\hline
\textbf{Febrile} & 969 (2.4\%) & 353 (1.5\%) & 616 (3.5\%) & $<$0.001 \\
\hline
\textbf{Hypotensive} & 651 (1.6\%) & 305 (1.3\%) & 346 (1.9\%) & $<$0.001 \\
\hline
\textbf{Gender} & & & & $<$0.001 \\
\hspace{0.5cm} Female & 22,105 (54\%) & 12,986 (55\%) & 9,119 (51\%) & \\
\hspace{0.5cm} Male & 19,092 (46\%) & 10,447 (45\%) & 8,645 (49\%) & \\
\hline
\textbf{Race} & & & & $<$0.001 \\
\hspace{0.5cm} White & 36,554 (89\%) & 20,665 (88\%) & 15,889 (89\%) & \\
\hspace{0.5cm} Black & 1,682 (4.1\%) & 974 (4.2\%) & 708 (4.0\%) & \\
\hspace{0.5cm} Asian & 1,217 (3.0\%) & 710 (3.0\%) & 507 (2.9\%) & \\
\hspace{0.5cm} Other & 737 (1.8\%) & 449 (1.9\%) & 288 (1.6\%) & \\
\hspace{0.5cm} Native & 538 (1.3\%) & 340 (1.5\%) & 198 (1.1\%) & \\
\hspace{0.5cm} Unknown & 469 (1.1\%) & 295 (1.3\%) & 174 (1.0\%) & \\
\hline
\hline
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

![Frequency of Sample Cases and Batches](../Manuscript/figures/fig1.png){width=40%, height=30%}

@@

Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED. In most EDs, however, physicians have discretion in picking their patients. In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm (Traub et al., 2016), which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Exhibit 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Wald Test Results}
    \begin{tabular}{lcc}
        \toprule
        Chief Complaints & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 1.37 & 0.106 \\
        Back or Flank Pain & 1.00 & 0.451 \\
        Chest Pain & 0.98 & 0.476 \\
        Extremity Complaints & 0.97 & 0.495 \\
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
        Gastrointestinal Issues & 0.98 & 0.480 \\
        Neurological Issue & 0.75 & 0.793 \\
        Shortness of Breath & 1.23 & 0.199 \\
        Skin Complaints & 1.05 & 0.388 \\
        Upper Respiratory Symptoms & 1.21 & 0.218 \\
        \midrule
        Emergency Severity & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 1.09 & 0.346 \\
        ESI 3, 4, or 5 & 1.24 & 0.196 \\
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\tiny
\item Table [blank] reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.
\end{tablenotes}
\end{table}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure 2 graphs the histogram of the leniency measure along the x-axis and the left y-axis. A local-linear regression of the fitted probability of batched ordering on tendency after residualizing is overlaid and displayed on the right y-axis. Table 2 presents the same ``first stage" in a regression table: being assigned to a 10 pp more lenient physician is associated with a 17 pp increase in the likelihood of being prescribed an opioid in the ED. The F-statistic is 25 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rarely prescribed diagnoses.

@@

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{tabular}{lccc}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Lab + Image Batching Tendency & 0.99*** & 1.02*** & 1.02*** \\
 & (0.01) & (0.01) & (0.01) \\
\\
F statistic (proj model) & 13206.68 & 6702.01 & 6655.86  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Image + Image Batching Tendency & 0.92*** & 0.94*** & 0.94*** \\
 & (0.02) & (0.04) & (0.04) \\
\\
F statistic (proj model) & 1645.35 & 577.97 & 577.09  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
\midrule
Num. obs. & 41197 & 41197 & 41197 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Notes:}  Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

### Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, if assignment of patients to ED physicians is random, conditional on seasonality, shift, and diagnosis (``conditional independence"). One can test for whether batch-likely physicians are routinely assigned patients with particular characteristics. Figure 4 shows evidence of balance on observables. The left panel is the output of a regression of Prescribed on observables including patient demographics, priority groups, and ED diagnoses. Unsurprisingly, many observables predict length of stay and 72 hour return. The right panel displays the coefficients from a regression of physician tendency instrument on the same set of observables. With so many coefficients, we fail to reject that the instrument violates conditional independence (the joint F-stat drops by two orders of magnitude, from 280 to 2.4). In fact, some of the variables that are significant in the second column go in the opposite direction of the decision to batch

The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch. In Appendix Figure [XXX], residualizing for only seasonality and shift (left panel) and also diagnosis (right panel) in equation (1). This lends support to our assumption that assignment in the ED is nearly random at the seasonality, shift  level.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section IIIB as well as various robustness checks in Section IIIE.

## Results

### Reduced-Form Results

@@

@@

\begin{table}[ht]
\centering
\caption{Regression Results by Chief Complaint} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & -0.72** & 0.03 \\ 
  2 & Upper Respiratory Symptoms & 0.426** & 0.02 \\ 
  3 & Abnormal Test Results & -0.067 & 0.84 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.423* & 0.10 \\ 
  5 & Shortness of Breath & -0.065 & 0.58 \\ 
  6 & Gastrointestinal Issues & 0.071 & 0.67 \\ 
  7 & Chest Pain & 0.107 & 0.68 \\ 
  8 & Dizziness/Lightheadedness/Syncope & -0.163 & 0.32 \\ 
  9 & Neurological Issue & -0.353 & 0.25 \\ 
  10 & Fatigue and Weakness & -0.112 & 0.49 \\ 
  11 & Back or Flank Pain & -0.602** & 0.02 \\ 
  12 & Abdominal Complaints & -0.105 & 0.13 \\ 
  13 & Fevers, Sweats or Chills & 0.01 & 0.95 \\ 
  14 & Skin Complaints & 1.442 & 0.35 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

---
title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization. A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity. Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes. The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

# Introduction {#sec:I}

@@

We start by evaluating the reduced-form effects of ED provider batch-ordering tendency on downstream patient outcomes and turnaround time. We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences. Being treated by a provider in the top decile of the tendency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].

Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes. This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization. It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.

The remainder of this paper is structured as follows. The next section describes the data source and outlines our baseline sample. The empirical strategy and its accompanying identifying assumptions are laid out in Section III. Section IV presents the results. Section V draws implications for batch-ordering policies. The last section concludes.

# Data and Definitions {#sec:II}

@@

In the context of diagnostic test ordering, we defined batching as placing multiple test orders within 5 minutes of one another. Sensitivity analyses on this cutoff point showed that our results are robust to this definition. Our study differentiated between two distinct batching strategies:

\begin{itemize}
\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.

\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
\end{itemize}

The distinction is necessary because laboratory tests may be processed in parallel with imaging tests, which have a much more limited effect on patient waiting time. However, a batch of multiple imaging tests requires that the patient is present for both tests. Thus, waiting times are much more likely to be impacted. Therefore, the specific tests that make up a batch may provide insight into how the physician trades off between waiting times and uncertainty, given the various complexity levels of patient encounters. Figure 1 shows the rate at which each batching type occurs for the top ten most frequently occurring chief complaint fields.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
  \tiny
  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
\end{figure}

@@

Table 2 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered individually in a batched or sequential way during an ED encounter from our baseline sample. About 39% of patient encounters involved diagnostic tests ordered as part of a batch, most of which (64%) included two tests.

Notable distinctions emerge when comparing the two cohorts. Patients subjected to batched test ordering exhibited a notably prolonged stay in the emergency department (ED) (p<0.001) and were more frequently assigned a lower Emergency Severity Index (ESI). Furthermore, this group skewed towards older demographics, with half of them falling into the 65+ age category. Patients in the batch-order group also displayed a higher likelihood of presenting symptoms at triage, including tachycardia (p<0.001), tachypnea (p<0.001), fever (p<0.001), and hypotension (p<0.001). Delving deeper into the ESI data unveils a higher concentration of patients with more critical conditions (ESI 1 and 2) within the batched group, in contrast to the sequential group (45.1% with ESI of 1 or 2 in the batched group compared to 27.5% with ESI of 1 or 2 in the sequential group). This finding suggests that factors related to patient complexity and severity (Saghafian et al., 2014) significantly influence physicians' decisions to opt for batch-ordering of diagnostic tests. However, it is important to note that our investigation revealed substantial variation among physicians in their choices between batching and sequencing for a given patient.

\begin{table}
\centering
\caption{Summary Statistics}
\begin{tabular}{p{3cm} *{4}{p{2.5cm}}}
\toprule
Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
\midrule
ESI & & & & $<0.001$ \\
1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
(Missing) & 27 & 16 & 11 & \\
\midrule
Age & & & & $<0.001$ \\
<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
\midrule
ED LOS (min) & & & & $<0.001$ \\
Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
\midrule
Disposition & & & & $<0.001$ \\
Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
\midrule
Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
\midrule
Number of Tests & & & & $<0.001$ \\
1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\midrule
Gender & & & & $<0.001$ \\
Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
\midrule
Race & & & & $0.002$ \\
White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

@@

Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED. In most EDs, however, physicians have discretion in picking their patients. In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm (Traub et al., 2016), which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Table 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Wald Test Results}
    \begin{tabular}{lcc}
        \toprule
        Chief Complaints & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 1.37 & 0.106 \\
        Back or Flank Pain & 1.00 & 0.451 \\
        Chest Pain & 0.98 & 0.476 \\
        Extremity Complaints & 0.97 & 0.495 \\
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
        Gastrointestinal Issues & 0.98 & 0.480 \\
        Neurological Issue & 0.75 & 0.793 \\
        Shortness of Breath & 1.23 & 0.199 \\
        Skin Complaints & 1.05 & 0.388 \\
        Upper Respiratory Symptoms & 1.21 & 0.218 \\
        \midrule
        Emergency Severity & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 1.09 & 0.346 \\
        ESI 3, 4, or 5 & 1.24 & 0.196 \\
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\tiny
\item Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.
\end{tablenotes}
\end{table}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure 2 graphs the histogram of batch-ordering frequency by physician for populat chief complaints, highlighting that the variation in batching differs systematically. Table 3 presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 10 pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 51 when all controls and fixed effects are included.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{threeparttable}
\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model)&7.32&52.03&50.93\\  
F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43300 & 43300 & 43300 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabularx}
\begin{tablenotes}
\tiny
\item Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}\tiny
\end{threeparttable}
\label{table:regression}
\end{table}

@@

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization. 
and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section 4.2 as well as various robustness checks in Section 4.4.

Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects. It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician. The literature leveraging the judges design typically performs two informal tests for its implications. The first one provides that the first stage should be weakly positive for all subsamples (Dobbie, Goldin, and Yang 2018). The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (Bhuller et al. 2020). Appendix Table 2 presents both of these tests in the two columns for various subsamples of interest. Finally, we check whether our main results hold using differential, mutually exclusive leniency measures (i.e., by complaint category) in Section 4.4.

# Results {#sec:4}

## Reduced-Form Results

@@

## Placebo Check

## Instrumental Variables Results

## Robustness

@@

:::
